Interval-Valued Quantum Probabilities: Contextuality and the Born Rule

### 2017-07-06 prop_letter.tex
done except abstract and intro

\begin{abstract}~
\amr{rewrite from scratch}
\begin{description}
\item [{Background}] In a world of bounded resource, any quantum measurement
can be only performed in finite precision.
\item [{Method}] We model the idea of the finite precision measurement
as finite number of intervals, and replace quantum real-valued probability
measures by interval-valued ones.
\item [{Results}] Quantum interval-valued probability measures provide
a unified framework to interpret Gleason's theorem and the Kochen-Specker
theorem as the limiting cases. For the cases between them, we found
some legitimate quantum interval-valued probability measures which
cannot be explained by the Born rule. While restricting on only commuting
projectors, the interval-valued probability measures become classical,
and these non-Born interval-valued probability measures disappear.
\item [{Conclusions}] Our research provides a new insight of finite precision
measurement and non-commutativity when formulating quantum theory.
Also, the Born rule may not be as nature as we used to believe in
the world of finite resource.
\end{description}
\end{abstract}

@@

In the classical setting, measurements done with limited resources and
poor precision may not uniquely identify the true state of a system
system. However, there is always at least one state that is consistent
with the imprecise
measurements\nb\cite{Shapley1971,GilboaSchmeidler1994,Grabisch2016}.

\begin{thm}\label{thm:Shapley}
  For every QIVPM\nb$\bar{\mu}:\events\rightarrow\mathscr{I}$, if a
  sub-event space\nb$\events'\subseteq\events$ commutes, then
  $\coreBorn\left(\bar{\mu},\events'\right)\ne\emptyset$.
\end{thm}

Gleason's theorem asserts any quantum real-valued probability measure
is consistent with a unique mixed state, and Shapley proved a quantum
IVPM must approximate some states for each classical sub-event
space. However, it may not approximate the same state in every
sub-event space. As the result, there may be no state consistent with
a QIVPM on all projectors\nb$\events$.

\begin{example}\label{ex:three-dimensional-three-value}
  Given a three dimensional Hilbert space with an orthonormal basis
  $\left\{ \ket{0},\ket{1},\ket{2}\right\} $.  Let
  $\mathscr{I}_{1}=\left\{ \necess,\imposs,\unknown\right\} $,
  $\ket{\ps}=\left(\ket{0}+\ket{1}\right)/\sqrt{2}$,
  $\ket{\ps'}=\left(\ket{0}+\ket{2}\right)/\sqrt{2}$, and
\begin{subequations}
\begin{eqnarray}
 &  & \begin{aligned}\imposs & =\bar{\mu}_{3}(\mathbb{0})=\bar{\mu}_{3}(\proj{0})\\
 & =\bar{\mu}_{3}(\proj{\ps})=\bar{\mu}_{3}(\proj{\ps'})\,,
\end{aligned}
\\
 &  & \begin{aligned}\necess & =\bar{\mu}_{3}(\mathbb{1})=\bar{\mu}_{3}(\mathbb{1}-\proj{0})\\
 & =\bar{\mu}_{3}(\mathbb{1}-\proj{\ps})=\bar{\mu}_{3}(\mathbb{1}-\proj{\ps'})\,,
\end{aligned}
\\
 &  & \unknown=\bar{\mu}_{3}(P)\qquad\textrm{otherwise.}
\end{eqnarray}
\end{subequations}
It is straightforward to verify that $\bar{\mu}_{3}$ is a QIVPM.

Let $S_{1}=\left\{ \proj{0},\proj{\ps}\right\} $. Because
$\bar{\mu}_{3}(\proj{0})=\bar{\mu}_{3}(\proj{\ps})=\imposs$, we have
$\mu_{\rho}^{\mathrm{B}}(\proj{0})=\mu_{\rho}^{\mathrm{B}}(\proj{\ps})=0$
for any mixed
state\nb$\rho\in\coreBorn\left(\bar{\mu}_{3},S_{1}\right)$.  Hence,
$\rho$ cannot be any state other than $\ket{2}$, the only pure state orthogonal to both
$\ket{0}$ and $\ket{\ps}$, and it is easy to verify $\proj{2}$ is actually
in $\coreBorn\left(\bar{\mu}_{3},S_{1}\right)$.

Similarly, let $S_{2}=\left\{ \proj{0},\proj{\ps'}\right\} $. We have
$\coreBorn\left(\bar{\mu}_{3},S_{2}\right)=\left\{ \proj{1}\right\} $.
Therefore,
\begin{equation}
\coreBorn\left(\bar{\mu}_{3},\events\right)\subseteq\coreBorn\left(\bar{\mu}_{3},S_{1}\right)\cap\coreBorn\left(\bar{\mu}_{3},S_{2}\right)=\emptyset
\end{equation}
and must be empty.
\end{example}

Although Gleason's theorem doesn't hold for every QIVPM, if a quantum
IVPM\nb$\bar{\mu}$ is induced by a totally mixed state\nb$\rho$
nicely, then whether another density
matrix\nb$\rho'\in\coreBorn\left(\bar{\mu},\events'\right)$ can be
estimated by $\left\Vert \rho-\rho'\right\Vert $, where a
state\nb$\rho$ is called totally mixed if
$0<\mu_{\rho}^{\mathrm{B}}\left(P\right)<1$ for any
$P\in\events\backslash\left\{ \mathbb{0},\mathbb{1}\right\} $, and the
the norm of a matrix\nb$A$ is defined by
$\left\Vert A\right\Vert
=\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{A}{\psi}\right|$\nb\citep{544199}.
For example, consider the intervals
\begin{eqnarray}
\widehat{\mathscr{I}}_{n} & = & \set{\left[\frac{k-1}{2n},\frac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\nonumber \\
 &  & \cup\left\{ \left[0,0\right],\left[1,1\right]\right\} 
\end{eqnarray}
which are overlapping intervals with length\nb$1/n$ except the intervals
near the ends. We can define an interval-valued function\nb$\widehat{\iota}_{n}:\left[0,1\right]\rightarrow\widehat{\mathscr{I}}_{n}$
by the following:
\begin{itemize}
\item when $x=0$ or $x=1$, $\widehat{\iota}_{n}\left(x\right)=\left[x,x\right]$;
\item when $0<x<1$, $\widehat{\iota}_{n}\left(x\right)=\left[\frac{k-1}{2n},\frac{k+1}{2n}\right]\cap\left[0,1\right]$,
where $k$ is defined by one of the following:
\begin{itemize}
\item $2\mid k$ and $k-0.5\le2nx\le k+0.5$;
\item $2\nmid k$ and $k-0.5<2nx<k+0.5$.
\end{itemize}
\end{itemize}
Then, for any totally mixed state\nb$\rho$ on a $d$-dimensional
Hilbert space, $\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}}$
is a QIVPM, and we have
\begin{itemize}
\item If $\left\Vert \rho-\rho'\right\Vert \le1/\left(4dn\right)$, $\rho'\in\coreBorn\left(\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}},\events\right)$.
\item If $\rho'\in\coreBorn\left(\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}},\events\right)$,
then $\left\Vert \rho-\rho'\right\Vert \le1/n$.
\end{itemize}
In summary, the states consistent with
$\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}}$ will be closer and
closer to $\rho$ as $n$ goes to infinity. In a limit case when the
intervals are infinitely precise, i.e.,
$\set{\left[x,x\right]}{x\in\left[0,1\right]}$, any QIVPM can be
recast as an real-valued one and has a unique state consistent with it
on every projector when $d\ge3$.

In general, this kind of inconsistency can be reduced if we increase
the measurement resource, that is, using finer intervals. We conjecture
by using finer and finer intervals, it would be more and more likely
that there is a state consistent with a QIVPM on all projectors\nb$\events$.

>>>>

\begin{abstract}~
\begin{description}
\item [{Background}] In a world of bounded resource, any quantum measurement
can be only performed in finite precision.
\item [{Method}] We model the idea of the finite precision measurement
as finite number of intervals, and replace quantum real-valued probability
measures by interval-valued ones.
\item [{Results}] Quantum interval-valued probability measures provide
a unified framework to interpret Gleason's theorem and the Kochen-Specker
theorem as the limiting cases. For the cases between them, we found
some legitimate quantum interval-valued probability measures which
cannot be explained by the Born rule. While restricting on only commuting
projectors, the interval-valued probability measures become classical,
and these non-Born interval-valued probability measures disappear.
\item [{Conclusions}] Our research provides a new insight of finite precision
measurement and non-commutativity when formulating quantum theory.
Also, the Born rule may not be as nature as we used to believe in
the world of finite resource.
\end{description}
\end{abstract}

@@

In the classical setting, measurements done with limited resources and
poor precision may not uniquely identify the true state of a
system. However, there is always at least one state that is consistent
with the imprecise
measurements~\cite{Shapley1971,GilboaSchmeidler1994,Grabisch2016}.

\begin{thm}\label{thm:Shapley}
  For a QIVPM\nb$\bar{\mu}:\events\rightarrow\mathscr{I}$, if a
  sub-event space\nb$\events'\subseteq\events$ commutes, then
  $\coreBorn\left(\bar{\mu},\events'\right)\ne\emptyset$.
\end{thm}

In the idealized situation of real-valued quantum probability
measures, Gleason's theorem asserts that every measure is determined
by a unique mixed state. The theorem above is weaker however: it
asserts that, for each classical sub-event subspace, a QIVPM must be
consistent with \emph{some} state. But the states consistent with
different subspaces may be incomparable as the following example
shows.

\begin{example}\label{ex:three-dimensional-three-value}
  Given a three dimensional Hilbert space with an orthonormal basis
  $\left\{ \ket{0},\ket{1},\ket{2}\right\} $.  Let
  $\mathscr{I}_{1}=\left\{ \necess,\imposs,\unknown\right\} $,
  $\ket{\ps}=\left(\ket{0}+\ket{1}\right)/\sqrt{2}$,
  $\ket{\ps'}=\left(\ket{0}+\ket{2}\right)/\sqrt{2}$, and
\begin{subequations}
\begin{eqnarray}
 &  & \begin{aligned}\imposs & =\bar{\mu}_{3}(\mathbb{0})=\bar{\mu}_{3}(\proj{0})\\
 & =\bar{\mu}_{3}(\proj{\ps})=\bar{\mu}_{3}(\proj{\ps'})\,,
\end{aligned}
\\
 &  & \begin{aligned}\necess & =\bar{\mu}_{3}(\mathbb{1})=\bar{\mu}_{3}(\mathbb{1}-\proj{0})\\
 & =\bar{\mu}_{3}(\mathbb{1}-\proj{\ps})=\bar{\mu}_{3}(\mathbb{1}-\proj{\ps'})\,,
\end{aligned}
\\
 &  & \unknown=\bar{\mu}_{3}(P)\qquad\textrm{otherwise.}
\end{eqnarray}
\end{subequations}
It is straightforward to verify that $\bar{\mu}_{3}$ is a QIVPM. Now
let $S_{1}=\left\{ \proj{0},\proj{\ps}\right\} $. Because
$\bar{\mu}_{3}(\proj{0})=\bar{\mu}_{3}(\proj{\ps})=\imposs$, we have
$\mu_{\rho}^{\mathrm{B}}(\proj{0})=\mu_{\rho}^{\mathrm{B}}(\proj{\ps})=0$
for any mixed
state\nb$\rho\in\coreBorn\left(\bar{\mu}_{3},S_{1}\right)$.  Hence,
$\rho$ cannot be any state other than $\ket{2}$, the only pure state
orthogonal to both $\ket{0}$ and $\ket{\ps}$, and it is easy to verify
$\proj{2}$ is actually in
$\coreBorn\left(\bar{\mu}_{3},S_{1}\right)$. Similarly, let
$S_{2}=\left\{ \proj{0},\proj{\ps'}\right\} $. We have
$\coreBorn\left(\bar{\mu}_{3},S_{2}\right)=\left\{ \proj{1}\right\} $.
Therefore,
\begin{equation}
\coreBorn\left(\bar{\mu}_{3},\events\right)\subseteq\coreBorn\left(\bar{\mu}_{3},S_{1}\right)\cap\coreBorn\left(\bar{\mu}_{3},S_{2}\right)=\emptyset
\end{equation}
Thus, there is no state that is consistent with the QIVPM on all
projectors\nb$\events$ invalidating Gleason's theorem.
\end{example}

Although Gleason's theorem does not hold for every QIVPM, we can show
that for a certain class of QIVPMs, the collection of states
consistent with the projectors get closer and closer to each other as
the intervals get narrower and narrower, eventually collapsing to a
single state when the intervals become infinitely precise real
numbers. To state this result, we introduce a few technical notions.

We call a state $\rho$ a \emph{totally mixed state} if
$0<\mu_{\rho}^{\mathrm{B}}\left(P\right)<1$ for any
$P\in\events\backslash\left\{ \mathbb{0},\mathbb{1}\right\} $. We also
define a notion of distance based on the following norm of a
matrix\nb$A$ as
$\left\Vert A\right\Vert
=\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{A}{\psi}\right|$\nb\citep{544199}. Now
consider the following collection of intervals:
\begin{eqnarray}
\widehat{\mathscr{I}}_{n} & = & \set{\left[\frac{k-1}{2n},\frac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\nonumber \\
 &  & \cup\left\{ \left[0,0\right],\left[1,1\right]\right\} 
\end{eqnarray}
which are overlapping intervals with length\nb$1/n$ except the intervals
near the ends.

A totally mixed state induces a QIVPM over the intervals above as
follows. We first define a map $\widehat{\iota}_{n}$ from the interval
$[0,1]$ to the collection of intervals above as follows:
\begin{itemize}
\item when $x=0$ or $x=1$, $\widehat{\iota}_{n}\left(x\right)=\left[x,x\right]$;
\item when $0<x<1$, $\widehat{\iota}_{n}\left(x\right)=\left[\frac{k-1}{2n},\frac{k+1}{2n}\right]\cap\left[0,1\right]$,
where $k$ is defined by one of the following:
\begin{itemize}
\item $2\mid k$ and $k-0.5\le2nx\le k+0.5$;
\item $2\nmid k$ and $k-0.5<2nx<k+0.5$.
\end{itemize}
\end{itemize}

It is now not difficult to verify that for any totally mixed
state\nb$\rho$ on a $d$-dimensional Hilbert space,
$\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}}$ is a
QIVPM. Furthemore, we have:
\begin{itemize}
\item If $\left\Vert \rho-\rho'\right\Vert \le1/\left(4dn\right)$, $\rho'\in\coreBorn\left(\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}},\events\right)$.
\item If $\rho'\in\coreBorn\left(\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}},\events\right)$,
then $\left\Vert \rho-\rho'\right\Vert \le1/n$.
\end{itemize}
In other words, the states consistent with this QIVMP
$\widehat{\iota}_{n}\circ\mu_{\rho}^{\mathrm{B}}$ will be closer and
closer to $\rho$ as $n$ goes to infinity. In the limiting case when
the intervals are infinitely precise, i.e.,
$\set{\left[x,x\right]}{x\in\left[0,1\right]}$, any QIVPM can be
recast as an real-valued one and will have a unique state consistent
with it on every projector when $d\ge3$.

% In general, this kind of inconsistency can be reduced if we increase
% the measurement resource, that is, using finer intervals. We conjecture
% by using finer and finer intervals, it would be more and more likely
% that there is a state consistent with a QIVPM on all projectors\nb$\events$.
% We can define an interval-valued function\nb
% :\left[0,1\right]\rightarrow\widehat{\mathscr{I}}_{n}$
% by the following:

% if a QIVPM\nb$\bar{\mu}$ is induced by a totally mixed
% state\nb$\rho$ nicely, then whether another density
% matrix\nb$\rho'\in\coreBorn\left(\bar{\mu},\events'\right)$ can be
% estimated by $\left\Vert \rho-\rho'\right\Vert
% $, where a
% state\nb$\rho$ is called totally mixed if
% $0<\mu_{\rho}^{\mathrm{B}}\left(P\right)<1$ for any
% $P\in\events\backslash\left\{ \mathbb{0},\mathbb{1}\right\}
% $, and the the norm of a matrix\nb$A$ is defined by $\left\Vert
%   A\right\Vert
% =\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{A}{\psi}\right|$\nb\citep{544199}.
% For example, consider the intervals
